Bootstrap learning via modular concept discovery

### 2014-06-08 old/drafts/ijcai2013/ijcai13.tex
Moved everything into an a new 'old' directory.

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{amsfonts}
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{float}
\usepackage{circuitikz}
\usepackage{graphicx}
\usepackage[font=small]{subcaption}
\usepackage[font=small]{caption}
\usepackage{alltt}
\usepackage{dashrule}
\usepackage{array}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning, patterns,backgrounds}
\usetikzlibrary{arrows,shapes.gates.logic.US,shapes.gates.logic.IEC,calc}

\usepackage{times}
\pdfinfo{
  /Title (Bootstrap learning via modular concept discovery)
  /Author (Eyal Dechter, Jon Malmaud, Ryan P. Adams, Joshua B. Tenenbaum)
}

@@

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Jon Malmaud \\
MIT \\ 
USA \\
malmaud@mit.edu
\And 
Ryan~P.  Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu
\And
Joshua~B. Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
}

@@

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer the learner effectively. How can such a learner ever
 get off the ground? A common intuition is that if the solutions to
 these problems share a common structure, and the learner can solve
 some simple problems by brute force, it should be able to extract
 useful components from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and Boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

\section{Introduction}
A central question for AI is how to create intelligent agents without
building in vast amounts of pre-specified knowledge that is tailored
to one domain or another. Any attempt to solve this problem must
contend with the fact that the less an agent knows about solving
problems in a domain, the worse its attempts are, and, therefore, the
less informative they are about how to solve subsequent problems.

For example, imagine you know nothing about electronics but are
supplied with a collection of circuit elements and asked to build a
useful device: your initial attempts will be uniformly
ineffective. If, on the other hand, you know the basics, you are far
more likely to generate devices that are somewhat successful and on
whose basis you might improve your skills.

Like many of the problems in AI, this one can be cast as a search
problem: the goal is to find a circuit wiring procedure that maximizes
certain criteria.  There are many search algorithms, but they typically
exploit local or topological structure in the search space. For
example, local search algorithms move from one candidate solution to a
neighboring one. Heuristic search techniques take advantage of
topological structure to prune off large parts of the solution space.

Domain specific knowledge provides this structure. In local search, it
defines the similarity metric between solutions; in heuristic search
it provides the heuristic. Here we ask how to bootstrap domain
specific knowledge and extract structure from the space of possible
solutions by attempting to solve multiple related problems in a given
domain at once. Our approach is motivated by the observation that many
interesting problems -- such as discovering scientific theories,
developing new technologies, learning to play a new instrument -- have
solutions that are natrually specified as programs.

Our proposed solution is thus inspired by the programmer who -- having
reused subroutines to solve similar simple tasks -- creates a function
library to help manage the complexity of larger projects. The
E.C. algorithm, which we introduce here, builds a library of reusable
program components and places a distribution over these, effectively
learning from simple tasks a search space structure that enables it to
solve more complex tasks.

@@

Concretely, the E.C. algorithm works as follows. It begins with a small
library of program primitives and a distribution over programs
synthesized from this library. In each iteration, it generates
programs in order of highest probability according to the library
learned in the previous iteration. From the discovered solutions, it
finds new reusable functions that compress these solutions, and it
re-estimates the distribution over programs by weighting these new
functions according to their frequencies. In using compression as a
guide to the choice of representation, we instantiate the idea that
good representations are those which minimize the description length
of typical elements in the domain.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is part of an
old tradition in AI. For instance, it is one of the central themes in
Herb Simon's 1969 ``The Sciences of the
Artificial''~\shortcite{simon1996sciences}. Rendell's 1985
paper~\shortcite{DBLP:conf/ijcai/Rendell85}, ``Substantial Constructive
Induction Using Layered Information Compression: Tractable Feature
Formation in Search,'' presented within a classical search paradigm
the idea that compression provides a heuristic for constructing good
representations. Within the Inductive Logic Programming (ILP)
literature, researchers have explored ``Constructive Induction,''
generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. Later, predicate invention
was also applied in the multitask
setting~\cite{DBLP:conf/ilp/KhanMP98}. We apply these ideas to
learning functional programs and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

Genetic programming
(GP)~\cite{DBLP:conf/foga/Koza92,DBLP:conf/ices/KozaBAK96} has
tackled the problem of program learning and has explored the notion
that reusable functions are helpful. This work relies, however, on
stochastic local search over the space of programs, and automatically
discovering useful subroutines is seen as a helpful heuristic. More
recently, Liang et al.~\shortcite{liang10programs} used a stochastic
grammar over combinatory logic to induce shared structure in
multi-task program induction problems. We find this latter
representation compelling and use it here, but we see both this and
genetic programming as tackling the question of how to benefit from
shared structure in situations where local search might be successful
on its own. Such problems rely on domain knowledge -- whether in the
form of a fitness function or a likelihood function -- to provide
local structure to the search space.

@@

The remainder of this paper will include two major sections. In the
first, we describe in detail the E.C. algorithm and design choices we
make in implementing it. We will then present results from two
experiments in the domains of symbolic regression and Boolean
function learning.

@@

We propose the \textbf{E.C. algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

The E.C. algorithm maintains a distribution $\mathcal{D}$ over
expressions in $\mathcal{L}$. At each iteration, it
\begin{enumerate}
\item \textbf{explores} the frontier -- enumerating the $N$ most probable expressions
  from the current distribution $\mathcal{D}$ -- and
\item \textbf{compresses} the hit tasks in the frontier to estimate a new
  distribution.
\end{enumerate}

@@

The basis combinators can themselves be expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction -- but in using the
combinatory logic we sequester uses of the abstraction operation
inside the combinators. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. Our representation thus becomes
variable-free. See \cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, Booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ \tau_1 \rightarrow \tau_2$ of
functions from source type $\tau_1$ to target type $\tau_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

\begin{figure}
%%  \centering
  \begin{subfigure}[Before]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow$ \\ $t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
    \caption{    \label{fig:clbintree}}
  \end{subfigure}
  \begin{subfigure}[After]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable expressions. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05,DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles,DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator OR it is a
left child program applied to a right child program (that is, a left
child AND a right child).

@@

  \vspace{2mm}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 4 ]
     \end{tikzpicture}
     \caption{(16)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 5 ]
     \end{tikzpicture}
     \caption{(25)}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 6 ]
     \end{tikzpicture}
     \caption{(36)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 7 ]
     \end{tikzpicture}
     \caption{(49)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 8 ]
     \end{tikzpicture}
     \caption{(64)}
  \end{subfigure}
\end{figure}
\captionsetup[subfigure]{labelformat=parens} 
\vspace{-.3cm}

\noindent The top row consists of one type of representation of an integer
squared, namely, the integer times itself. The bottom row contains
another type of representation in which the function $f(x)=x^2$,
represented in combinatory logic as $S * I$, is applied to the
integer. The top and bottom row of each expression evaluate to the
same integer. If we consider only the first column, the top tree has 4
unique subtrees and the bottom has 7. But each column adds 3 unique
subtrees to the top row and only 2 to the bottom. So by the fifth
column the top row has more unique subtrees than the bottom row. The
increased compressibility of the bottom row is due to the fact that
the squaring operation has been factored out as an abstraction. This
is an example of how this compressibility metric favors
abstraction when multiple expressions are being considered at once.

@@

Given a set of chosen solution expressions, which we will call the
\emph{solution set}, want to re-estimate our stochastic grammar. Our
inspiration for doing this is the Neville-Manning algorithm for
grammar-based compression~\cite{nevill1997identifying}. The idea of
that compression algorithm is to compress any string by creating a
grammar for that string such that a) no two symbols appear more than
once in the grammar and b) every rule is used more than once. From a
set of expressions, we generate a grammar according to these criteria.

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if it returns the
same value when applied to $i \in 0, \dots, 9$ that $f$ does. In the
experiment shown here, the set of problems corresponds to the set of
all polynomials with degree less than or equal to 2. In the initial
library, we include the basic combinators I, S, B, C and four
arithmetic primitives $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks (dashed pink line). \label{fig:symregLearning}}
\end{figure}
\begin{figure}[ht]
\includegraphics[width=1.07\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at frontier size of $10000$ for different task
  sets.} 
\label{fig:symregCurricula}

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we
vary the frontier size. Changing the frontier size changes the number
of tasks the algorithm identifies correctly over the course of 15
algorithm iterations (all experiments in this work were run for 15
iterations). As a baseline, we enumerated $10000*15=150000$
expressions from the initial grammar. This is the total number of
expressions that a run of the algorithm sees if it has a frontier size
of $10000$ and runs for $15$ iterations. Thus, if the benefit accruing
to the runs with larger frontier sizes is simply due to an increase in
the number of expressions seen, rather than increased learning, we
should see similarly good performance from this baseline run. In fact,
however, the baseline run only hits $27$ of the $1000$ tasks ($3\%$),
whereas our algorithm nears $100\%$ for a frontier of $10000$.

What does the E.C. algorithm learn in this task?  Inspecting the top
weighted primitives of the final grammars, we find many incrementers
(e.g. (+1), (+2), etc.), several versions of expressions that
translate to functions like $x*f(x)$ and $f(x*f(x))$, and combinations
of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally, we see
expressions that apply functions to themselves, building up complex
functions with relatively few unique primitives.

To what degree is ``conceptual bootstrapping'' responsible for the
improved performance. That is, to what extent does the presence of
simple problems account for the ability to learn complex functions?
One hypothesis is that to succeed on a set of tasks, the set must
contain a ``learnable'' curriculum, a set of tasks that serve as
stepping stones from an impoverished representation to a rich one. We
can test this hypothesis by varying the set of tasks to which the E.C.
algorithm is exposed. If it is true, then we should see a nonlinear
response to reducing the number of easy tasks, as the curriculum
changes from being learnable to being unlearnable.

>>>>

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{float}
\usepackage{circuitikz}
\usepackage{graphicx}
\usepackage[font=small]{subcaption}
\usepackage[font=small]{caption}
\usepackage{alltt}
\usepackage{dashrule}
\usepackage{array}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning, patterns,backgrounds}
\usetikzlibrary{arrows,shapes.gates.logic.US,shapes.gates.logic.IEC,calc}

@@

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Jon Malmaud \\
MIT \\ 
USA \\
malmaud@mit.edu
\And 
Ryan P.  Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu
\And
Joshua B. Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
}

@@

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

\section{Introduction}
Many of the problems we want to solve in AI are best cast as search
problems. There are many search algorithms, but they typically exploit
local or topological structure in the search space. For example,
convex optimization and gradient based search methods take
advantage of local smoothness to move in good directions. Local search
algorithms move from one candidate solution to a neighboring
one. Heuristic search techniques take advantage of topological
structure to prune off large parts of the solution space.

However, many of the problems we want solve -- discovering scientific
theories, developing new technologies, figuring out how to bake -- do
not come with such a structured search space. This is particularly
true of problems, such as those mentioned, whose solutions can be
naturally framed as programs. Here we ask how to extract structure
from the space of programs by attempting to solve multiple related
problems in a given domain at once.

We take as inspiration the programmer who -- having reused several
subroutines to solve different but similar simple tasks -- creates a
function library to help manage the complexity of larger projects. The
EC algorithm, which we introduce here, builds a library of reusable
program components and places a distribution over these, effectively
learning from simple tasks a search space structure that enables it
to solve more complex tasks.

@@

Concretely, the EC algorithm works as follows. It begins with a small
library of program primitives and a distribution over programs
synthesized from this library. In each iteration, it generates
programs in order of highest probability according to the library
learned in the previous iteration. From the discovered solutions, it
finds new reusable functions that compress these solutions, and it
re-estimates the distribution over programs by weighting these new
functions according to their frequencies. In using compression as a
guide to the choice of representation, we instantiate the idea that
good representations are those which minimize the description length
of typical elements in the domain.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is part of an
old tradition in AI. For instance, it is one of the central themes in
Herb Simon's 1969 ``The Sciences of the
Artificial''~\cite{simon1996sciences}. Rendell's 1985
paper~\cite{DBLP:conf/ijcai/Rendell85}, ``Substantial Constructive
Induction Using Layered Information Compression: Tractable Feature
Formation in Search,'' presented within a classical search paradigm
the idea that compression provides a heuristic for constructing good
representations. Within the Inductive Logic Programming (ILP)
literature, researchers have explored ``Constructive Induction,''
generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. Later, predicate invention
was also applied in the multitask
setting~\cite{DBLP:conf/ilp/KhanMP98}. We apply these ideas to
learning functional programs, and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

Genetic programming
(GP)~\cite{DBLP:conf/foga/Koza92}~\cite{DBLP:conf/ices/KozaBAK96} has
tackled the problem of program learning and has explored the notion
that reusable functions are helpful. This work relies, however, on
stochastic local search over the space of programs, and automatically
discovering useful subroutines is seen as a helpful heuristic. More
recently, Liang et al.~\cite{liang10programs} used a stochastic
grammar over combinatory logic to induce shared structure in
multi-task program induction problems. We find this latter
representation compelling and use it here, but we see both this and
genetic programming as tackling the question of how to benefit from
shared structure in situations where local search might be successful
on its own. Such problems rely on domain knowledge -- whether in the
form of a fitness function or a likelihood function -- to provide
local structure to the search space.

@@

The remainder of this paper will include two major sections. In the
first, we describe in detail the EC algorithm and design choices we
make in implementing it. We will then present results from two
experiments in the domains of symbolic regression and Boolean
function learning.

@@

We propose the \textbf{EC algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

The EC algorithm maintains a distribution $\mathcal{D}$ over
expressions in $\mathcal{L}$. At each iteration, it
\begin{enumerate}
\item \textbf{explores} the frontier -- enumerating the $N$ most probable expressions
  from the current distribution $\mathcal{D}$ -- and
\item \textbf{compresses} the hit tasks in the frontier to estimate a new
  distribution.
\end{enumerate}

@@

The basis combinators can themselves be expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction -- but in using the
combinatory logic we sequester uses of the abstraction operation
inside the combinators. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. Our representation thus becomes
variable-free. See Liang et al.~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ \tau_1 \rightarrow \tau_2$ of
functions from source type $\tau_1$ to target type $\tau_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

\begin{figure}
%%  \centering
  \begin{subfigure}[Before]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      }
    \end{tikzpicture}
    \caption{    \label{fig:clbintree}}
  \end{subfigure}
  \begin{subfigure}[After]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable expressions. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator OR it is a
left child program applied to a right child program (that is, a left
child AND a right child).

@@

  \vspace{2mm}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 4 ]
     \end{tikzpicture}
     \caption{(16)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 5 ]
     \end{tikzpicture}
     \caption{(25)}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 6 ]
     \end{tikzpicture}
     \caption{(36)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 7 ]
     \end{tikzpicture}
     \caption{(49)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 8 ]
     \end{tikzpicture}
     \caption{(64)}
  \end{subfigure}
\end{figure}
\captionsetup[subfigure]{labelformat=parens} 
\vspace{-.3cm}
The top row consists of one type of representation of an integer
squared, namely, the integer times itself. The bottom row contains
another type of representation in which the function $f(x)=x^2$,
represented in combinatory logic as $S * I$, is applied to the
integer. The top and bottom row of each expression evaluate to the
same integer. If we consider only the first column, the top tree has 4
unique subtrees and the bottom has 7. But each column adds 3 unique
subtrees to the top row and only 2 to the bottom. So by the fifth
column the top row has more unique subtrees than the bottom row. The
increased compressibility of the bottom row is due to the fact that
the squaring operation has been factored out as an abstraction. This
is an example of how this compressibility metric favors
abstraction when multiple expressions are being considered at once.

@@

Given a set of chosen solution expressions, which we will call the
\emph{solution set}, want to re-estimate our stochastic grammar. Our
inspiration for doing this is the Neville-Manning algorithm for
grammar-based compression~\cite{nevill1997identifying}. The idea of
that compression algorithm is to compress any string by creating a
grammar for that string such that a) no two symbols appear more than
once in the grammar and b) every rule is used more than once. From a
set of expressions we generate a grammar according to these criteria.

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if it returns the
same value when applied $i \in 0, \dots, 9$ that $f$ does. In the
experiment shown here, the set of problems corresponds to the set of
all polynomials with degree less than or equal to 2. In the initial
library, we include the basic combinators I, S, B, C and four
arithmetic primitives, $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\includegraphics[width=.8\linewidth]{figures/symreg3frontiers.pdf}
\vspace{-.2cm}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\vspace{-.3cm}
\end{figure}
\begin{figure}[ht]
\centering
\includegraphics[width=.8\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at frontier size of $10000$ for different task
  sets.} 
\label{fig:symregCurricula}
\vspace{-.5cm}
\end{figure}

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we
vary the frontier size.Changing the frontier size changes the number
of tasks the algorithm identifies correctly over the course of 15
algorithm iterations (all experiments in this work were run for 15
iterations). As a baseline, we enumerated $10000*15=150000$
expressions from the initial grammar. This is the total number of
expressions that a run of the algorithm sees if it has a frontier size
of $10000$ and runs for $15$ iterations. Thus, if the benefit accruing
to the runs with larger frontier sizes is due to simply an increase in
the number of expressions seen, rather than increased learning, we should
see similarly good performance from this baseline run. In fact,
however, the baseline run only hits $27$ of the $1000$ tasks ($3\%$),
whereas our algorithm nears $100\%$ for a frontier of $10000$.

What does the EC algorithm learn in this task?  Inspecting the top
weighted primitives of the final grammars, we find many incrementers
(e.g. (+1), (+2), etc.), several versions of expressions that
translate to functions like $x*f(x)$ and $f(x*f(x))$, and combinations
of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally, we see
expressions that apply functions to themselves, building up complex
functions with relatively few unique primitives.

To what degree is ``conceptual bootstrapping'' responsible for the
improved performance. That is, to what extent does the presence of
simple problems account for the ability to learn complex functions?
One hypothesis is that to succeed on a set of tasks, the set must
contain a ``learnable'' curriculum, a set of tasks that serve as
stepping stones from an impoverished representation to a rich one. We
can test this hypothesis by varying the set of tasks to which the EC
algorithm is exposed. If it is true, then we should see a nonlinear
response to reducing the number of easy tasks, as the curriculum
changes from being learnable to being unlearnable.
